\bf Distortion corrected kernel density estimate on Riemannian manifolds

### 2023-06-01 jcgs/03-chap3-kderm.Rmd
Add jcgs version for kderm, the latest version

---
title: "Distortion corrected kernel density estimate on Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: The University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

\newpage

@@

Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Chen2017-dw and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq]. For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$. The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da and @Berenfeld2022-fd. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$. Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx and @Cheng2021-dh. One could easily apply KDE to the output vectors $\pmb{y}$. However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is 
$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,$$
where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$. By a change of variables, this would imply the following density for $\pmb{y}$,
\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
\end{equation}
where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$. It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant. Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g. ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.  Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].

@@

<!--

OLD:

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of $\mathbb{R}^d$. Applications of kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac], prime prediction via Twitter messages [@Gerber2014-tq], and so on. However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than $\mathbb{R}^d$, kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted. @Pelletier2005-vu propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known. Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold. Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than $\mathbb{R}^d$. Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function. This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.

Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data. It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-dh], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data. In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies. To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut]. 
There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details]. 
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews]. The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm. In this paper, we focus on the multivariate kernel density estimation.

Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed. 
The bandwidth is varied depending on either the location of the sample points [sample smoothing estimator; @Terrell1992-ut] or that of the estimated points [balloon estimator; @Terrell1992-ut]. In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the *variable/adaptive kernel density estimation* [VKDE; Section 6.6 of @Scott2015-vl] unless otherwise stated.
However, these kernel density estimators are based on random samples in the Euclidean space.

For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points. The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature [@Henry2009-ll]. 
However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space. Therefore, the kernel density estimator from @Pelletier2005-vu is not applicable because the geodesic distance and the volume density function are unknown.
This is when we introduce manifold learning to reduce the input data dimension. For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-dh), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold [@Goldberg2008-co]. To tackle this problem and measure the distortion incurred in manifold learning,
@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding. Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others. The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.

By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding. Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point. Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information. Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold. These density estimates are useful in many other areas, including classification, clustering and anomaly detection. Similar to @Cheng2021-dh, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds. Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection. \autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space. Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.

-->

@@

In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample. Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case. For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$. We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\pmb{y}\|)=K(0).
\end{align}
Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation. For instance, condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise. In our empirical section, more general kernel functions can be, and are, employed.

@@

<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density. To make this clear consider the simpler case computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term. A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density on the manifold itself.  Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density in its own. This motivates a kernel desnity estimate that corrects for the distortion induced by $\psi$.-->

@@

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters EMBED }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
  $$
    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
  $$
  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;

  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
  $$
    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
  $$
  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.

@@

```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="Underlying data for the Gaussian mixture model of four kenels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$. The colors indicate the true density of the data when they are mapped via the twin peaks function. Lower density points in darker colors are scattered both in the outer and center areas.", echo = FALSE}
knitr::include_graphics("figures/truedensity_twinpeaks_dc_labv.png")
```

@@

```{r twinpeaks3d, fig.align = 'center', out.width = "80%", fig.cap="(ref:twinpeaks3dcaption)", echo = FALSE}
knitr::include_graphics("figures/scatterplot3d_twinpeaks_dc.png")
```

@@

```{r tpoutliers, fig.align = 'center', out.width = "95%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row. The top 20 outliers, hightlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel) and KDE (right panel). DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
```

@@

```{r tpcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_Twin Peak2000_5ml_r0_5.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>% 
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data. Distortion corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>%
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>% 
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4]) %>%
  column_spec(6, bold = max.match[,5])
```

@@

<!-- ## Irish smart meter dataset -->

@@

<!-- kernel types discussion -->

@@

<!-- \clearpage -->
<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
<!-- %\pagenumbering{arabic} -->
<!-- %\def\thepage{A\arabic{page}} -->

\appendix

# Appendix: Notions about Riemannian geometry {#riemgeo}

>>>>

@@

@@

Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Scott2015-vl and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq]. For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$. The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping $\psi:M\rightarrow \mathbb{R}^d$ that take points on the manifold $\pmb{p}$ to $d$-dimensional vectors $\pmb{y}$. Details of these algorithms are summarized in @Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh. One could easily apply KDE to the output vectors $\pmb{y}$. However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is 
$$Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,$$
where $f$ is the probability density function on the manifold, $\pmb{p}$ is a point on the manifold, $\mathcal{A}$ is some region on the manifold and $dvol_M$ is a volume form on the manifold $M$. By a change of variables, this would imply the following density for $\pmb{y}$,
\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
\end{equation}
where $\psi(\mathcal{A})$ is the image of $\mathcal{A}$ under $\psi$ and $\pmb{H}(\pmb{y})$ is the Riemannian metric expressed in local coordinates given by the mapping $\psi$. It can be seen that the density in terms of $\pmb{y}$, differs from the density on the manifold due to the term $|\pmb{H}(\pmb{y})|^{1/2}$ which can be considered roughly analogous to the Jacobian determinant. Since $\psi$ will typically be a non-linear function, the value of $|\pmb{H}(\pmb{y})|^{1/2}$ will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different $\psi$, meaning that if KDE is applied to the output $\pmb{y}$ coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of $|\pmb{H}(\pmb{y})|^{1/2}$ for this purpose we use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g. ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of $\pmb{y}$ to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space $\pmb{y}$ is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm 'squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data.  Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic [see Section 6.6 of @Scott2015-vl for a summary].

@@

@@

In the following, we denote $M$ as the $d$-dimensional manifold from which our data are sampled. Points on this manifold are denoted $\pmb{p}$ in general, with $\pmb{p}_1,\dots,\pmb{p}_n$ denoting the observed sample. Often $\pmb{p}_i$ will be high-dimensional vectors such that $\pmb{p}_i\in\mathbb{R}^m$ with $m\gg d$, however, this need not be the case. For instance, $\pmb{p}_i$ may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each $\pmb{p}_i$ only require some sense of distance between the 'input' points, $d(\pmb{p}_i,\pmb{p}_j)$, such that we can apply dimension reduction algorithms to obtain an `output' embedding, $\pmb{y}_1,\dots,\pmb{y}_n$, where $\pmb{y}_i\in\mathbb{R}^d$. We will denote this embedding as $\pmb{y}_i=\psi(\pmb{p}_i)$. Finally, we denote by $\lambda$ the Lebesgue measure of $\mathbb{R}^d$, letting $\|\cdot\|$ be the usual Euclidean norm and following @Pelletier2005-vu, we make the following assumptions about the kernel function $K:\mathbb{R}_+\rightarrow\mathbb{R}$,
\begin{equation}
\begin{aligned}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=1;
(ii) \int_{\mathbb{R}^{d}} \pmb{y} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\pmb{y}\|^{2} K(\|\pmb{y}\|) \mathrm{d} \lambda(\pmb{y})<\infty; \\
& (iv) \supp K=[0; 1]; 
(v) \sup K(\|\pmb{y}\|)=K(0).
\end{aligned}
\end{equation}

Note that these conditions are different from (and in some cases stricter than) those normally used for kernel density estimation. For instance, condition (iv) requires the support of the kernel to be bounded. The reasons for this will become clearer when we discuss the manifold setting in more detail. Also, for illustration purposes, in this section, we pay particular attention to the uniform kernel for which $K(z)$ equals one if $0\leq z\leq1$ and zero otherwise. In our empirical section, more general kernel functions can be, and are, employed.

@@

<!--As a non-linear transformation, any dimension reduction algorithm `distorts' the density. To make this clear consider the simpler case of computing the density after a change of variables $\psi:\mathbb{R}^d\rightarrow\mathbb{R}^d$, which involves a Jacobian term. A similar notion applies to a manifold embedding so that the density of the output vectors $\pmb{y}_i$ differs from the density of the manifold itself.  Furthermore, standard kernel density estimates applied directly on the output embedding will be sensitive to the choice of dimension reduction algorithm since each different algorithm will distort the density on its own. This motivates a kernel density estimate that corrects for the distortion induced by $\psi$.-->

@@

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm EMBED }
  \OptParameter{ manifold learning parameters in EMBED }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \STATE Obtain the $d\times d\times N$ tensor $\pmb{\tilde{H}}$ by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with elements in $\pmb{\tilde{H}}$ found by
  $$
    \pmb{\tilde{H}}^{lk \cdot} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^l \odot \pmb{y}^k\right) - \pmb{y}^l \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^k\right) - \pmb{y}^k \odot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^l\right)\right],
  $$
  where $l,k=1,\dots,d$, $\pmb{\tilde{H}}^{lki}$ is the element in row $l$ and column $k$ of the inverse Riemannian corresponding to observation $i$, and the $\odot$ calculation is the element-wise product between two vectors;

  \STATE For each embedding data point $\pmb{y}_i, i=1,\dots,N$, calculate its $d\times d$ Riemannian metric $\pmb{H}_i$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}^{\cdot\cdot i}}$ below 
  $$
    \pmb{H}_i = \pmb{U} \pmb{\Lambda}_d^{-1} \pmb{U}^\prime,
  $$
  where $[\pmb{U}, \pmb{\Lambda}]$ is the eigendecomposition of the $d\times d$ matrix $\tilde{\pmb{H}^{\cdot\cdot i}}$, $\pmb{U}$ is the matrix of column eigenvectors ordered by the eigenvalues in descending order, and $\pmb{\Lambda}_d$ is a diagonal matrix containing the top $d$ largest eigenvalues.

@@

```{r metadensity, fig.align = 'center', out.width = "80%", fig.cap="Underlying data for the Gaussian mixture model of four kernels with means $(0.25, 0.25), (0.25, 0.75), (0.75, 0.25), (0.75, 0.75)$ and the same variance-covariance matrix $diag(0.016, 0.016)$. The colors indicate the true density of the data when they are mapped via the twin peaks function. Lower density points in darker colors are scattered both in the outer and center areas.", echo = FALSE}
knitr::include_graphics("figures/truedensity_twinpeaks_dc_labv.png")
```

@@

```{r twinpeaks3d, fig.align = 'center', out.width = "75%", fig.cap="(ref:twinpeaks3dcaption)", echo = FALSE}
knitr::include_graphics("figures/scatterplot3d_twinpeaks_dc.png")
```

@@

```{r tpoutliers, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row. The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel). DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
```

@@

```{r tpcors, echo=FALSE, message=FALSE, eval=T}
load("figures/CorrelationTable_Twin Peak2000_5ml_r0_5.rda")
col.max <- apply(cors, 2, max)
max.match <- rep(col.max, each = 2) == cors
data.frame(cors) %>% 
  mutate_if(is.numeric, format, digits = 3, nsmall = 3) %>% 
  kableExtra::kbl(caption = "Correlation between true density ranking and estimated density ranking for different manifold learning embeddings of the twin peak data. Distortion-corrected KDE outperforms for all dimension reduction algorithms and gives the higher rank correlation to the output of t-SNE and UMAP.", booktabs = TRUE, digits = 3, escape = FALSE) %>%
  # kable_styling(latex_options = "scale_down") %>%
  kable_paper(full_width = FALSE) %>%
  # column_spec(1, width = "6cm") %>% 
  column_spec(2, bold = max.match[,1]) %>%
  column_spec(3, bold = max.match[,2]) %>%
  column_spec(4, bold = max.match[,3]) %>%
  column_spec(5, bold = max.match[,4]) %>%
  column_spec(6, bold = max.match[,5])
```

@@

<!-- # Irish smart meter dataset -->

@@

@@
